---
title: "Cyclistic Bike-share Analysis"
author: "Momoh Alvin"
date: "2022-07-14"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## [Business task]{.underline}

How do annual members and casual riders use Cyclistic bikes differently?

## [Data source used]{.underline}

Cyclistic's historical trip data was used to analyze and identify trends.[Click](https://divvy-tripdata.s3.amazonaws.com/index.html){.uri} for the previous 12 months of Cyclistic trip data. (Note: The datasets have a different name because Cyclistic is a fictional company. The data has been made available by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement){.uri}. This public data was used to explore how different customer types are using Cyclistic bikes.

## [Cyclistic Trip Data Cleaning]{.underline}

#### Google Sheets

Google Sheets was used to get a quick view of the entire Cyclistic Data set, to check if rows with missing values are present, to check for duplicate rows, then checked for other errors and Understanding the data.

#### Data cleaning with python

``` python
import pandas as pd
```

``` python
file_names = ["202101-divvy-tripdata.csv", "202102-divvy-tripdata.csv", "202103-divvy-tripdata.csv", "202104-divvy-tripdata.csv", "202105-divvy-tripdata.csv", "202106-divvy-tripdata.csv", "202107-divvy-tripdata.csv", "202108-divvy-tripdata.csv", "202109-divvy-tripdata.csv", "202110-divvy-tripdata.csv", "202111-divvy-tripdata.csv", "202112-divvy-tripdata.csv"]
```

``` python
a = 0 
num = "01"
for names in range(len(file_names)):
  # file name position in file_names
  name = file_names[a]

  # load data
  df = pd.read_csv(name)

  # drop rows with missing values (since there are no specifics on how to treat missing values)
  df = df.dropna()
  
#Date Manipulation

  #convert datatype to string
  bb = df[["started_at", "ended_at"]].astype("string")

  # temprary variables for string manipulation
  f4 = bb["started_at"].str.slice(stop = 4)
  n4 = bb["started_at"].str.slice(start = 4, stop = 8)
  l2 = bb["started_at"].str.slice(start = 8, stop = 10)
  rs = bb["started_at"].str.slice(start = 11)

  f41 = bb["ended_at"].str.slice(stop = 4)
  n41 = bb["ended_at"].str.slice(start = 4, stop = 8)
  l21 = bb["ended_at"].str.slice(start = 8, stop = 10)
  rs1 = bb["ended_at"].str.slice(start = 11)

  bn = l2 + n4 + f4
  bn1 = l21 + n41 + f41

  #convert columns to datetime
  df["started_at"] = pd.to_datetime(bn) + pd.to_timedelta(rs)
  df["ended_at"] = pd.to_datetime(bn1) + pd.to_timedelta(rs1)

  # get the amount of time bikes were used (duration of use)
  df["duration_min"] = df["ended_at"] - df["started_at"]

  df = df.sort_values("duration_min")


  #convert time to int
  df["duration_min"] = df["duration_min"].dt.total_seconds().astype(int)

  #filter data to get rid of abnormalities
  #(some days had bike usage above 24 hours. Since there is no data
  # how those should be dealt with, days above  24 hours where filtere
  #and droped)
  df = df[(df["duration_min"] > 60) & (df["duration_min"] < 86340)]

  #duration_min  to minutes
  df["duration_min"] = df["duration_min"] / 60

  # created day of week column
  df["week_day"] = df["ended_at"].dt.day_name()

  #data to be exported
  df = df[["started_at", "ended_at", "member_casual", "duration_min", "week_day"]]

  # export adjusted file
  df.to_csv("Cycl_trip_data_2021_" + num + ".csv", index = False #to adjust each file name)

  # incrementing num and a
  num = str(int(num) + 1)
  a += 1 
```

## [Summary of data]{.underline}

#### summary in BigQuery

##### Combining all data

``` sql
SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.ccyl\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_2\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_3\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_4\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_5\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_6\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_7\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_8\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_9\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_10\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_11\`

UNION ALL

SELECT \*

FROM \`cyclistic-345018.cln_cyl_data.cyl_12\`
```

##### Data grouped into weeks

``` sql
SELECT  
member_casual,
ryd_week,
ROUND(AVG(duration_in_min), 2) AS avg_dur_week,
ROUND(SUM (duration_in_min), 2) AS total_dur_wkly
FROM `cyclistic-345018.cln_cyl_data.cln_cyl_f1`
GROUP BY member_casual, ryd_week
ORDER BY ryd_week
```

##### Data grouped into months

``` sql
SELECT  
ryd_month,
member_casual,
ROUND(AVG(duration_in_min), 2) AS avg_dur_mnt,
ROUND(SUM (duration_in_min), 2) AS total_dur_mnt
FROM `cyclistic-345018.cln_cyl_data.cln_cyl_f1`
GROUP BY member_casual, ryd_month, ryd_mnt_num
ORDER BY ryd_mnt_num
```

##### Grouping by Subscription

``` sql
SELECT  
member_casual,
ROUND(AVG(duration_in_min), 2) AS avg_du_yr,
ROUND(SUM (duration_in_min), 2) AS total_dur_yr
FROM `cyclistic-345018.cln_cyl_data.cln_cyl_f1`
GROUP BY member_casual
```

## [visualizations and key findings]{.underline}

#### Importing Cyclistic Data from Google Sheets

Weekly Data (wkly_df), Monthly Data(mntly_df), Year's Data(yrs_df)

```{r}
wkly_data <- "https://docs.google.com/spreadsheets/d/e/2PACX-1vTcUdnCUI1phbae7bvjsKS8gLzloQqBtFJBPjoQiCxvV1YQ_igsJBoaZh9N1um9s-rS90dXTFO6oxIk/pub?output=csv"
mntly_data <- "https://docs.google.com/spreadsheets/d/e/2PACX-1vT_Z-wpKRvAD4UPwEvvopa4ioujaWVMr5vxdewZZ4uDtVUuze-oAuf736n04AuJirK-BiEkDgXpTpfd/pub?output=csv"

years_data <- "https://docs.google.com/spreadsheets/d/e/2PACX-1vSElKjjGf55ByzlNC1kv05mxNYCOaqJuzxqLQT7etMZTmtti-2umr0h344ws1IsfcUvHUJ2CJ7jNPOt/pub?output=csv"
```

```{r}
wk_df <- read.csv(url(wkly_data))
```

```{r}
library(magrittr)
library(dplyr)
library(stringr) 
library(ggplot2)


# to create a column that adds "wk"(week) to ride_week 
wk_df["rd_week"] <- sub("^", "wk", as.character(wk_df$ryd_week))

# to create a column that renames member_casual
wk_df$subscription <- wk_df$member_casual%>%str_replace_all(c("member" = "annual subs", "casual" = "casual riders"))

# to create a column that reduces the values shown on the x axis 
wk_df$time <- round((wk_df$total_dur_wkly/ 10000), 2)



# To split average duration into annual memberships and  casual riders to get their individual min and max duration 
member_df <- wk_df%>%filter(member_casual == "member")
MAXm <- max(member_df %>% select(avg_dur_week))
MINm <- min(member_df %>% select(avg_dur_week))


casual_df <- wk_df%>%filter(member_casual == "casual")
MAXc <- max(casual_df %>% select(avg_dur_week))
MINc <- min(casual_df %>% select(avg_dur_week))
wka_df <- wk_df%>%select(member_casual, ryd_week, avg_dur_week, rd_week, subscription, time)


```

### Weekly Data Set

`head(wka_df, 10)`

```{r, width= 8}
head(wka_df,  10)
```

| member_casual | Member/Casual | Ride Week | Average Duration | Ride Week (adjusted) | Subscription  |
|:-----------|:-----------|:-----------|:-----------|:-----------|:-----------|
| 1             | casual        | 0         | 36.56            | wk0                  | casual riders |
| 2             | member        | 0         | 14.25            | wk0                  | annual subs   |
| 3             | member        | 1         | 14.07            | wk1                  | annual subs   |
| 4             | casual        | 1         | 28.68            | wk1                  | casual riders |
| 5             | member        | 2         | 11.70            | wk2                  | annual subs   |
| 6             | casual        | 2         | 22.02            | wk2                  | casual riders |
| 7             | member        | 3         | 11.75            | wk3                  | annual subs   |
| 8             | casual        | 3         | 21.06            | wk3                  | casual riders |
| 9             | member        | 4         | 12.38            | wk4                  | annual subs   |
| 10            | casual        | 4         | 20.56            | wk4                  | casual riders |

: Cyclistic weekly bike usage

### Cyclistic Weekly Average Bike Usage(WAB) (min)

```{r, fig.width = 8, fig.height=4}
# to display plots
WAB <- 
  ggplot(wk_df, aes(x = ryd_week, y = avg_dur_week, color = subscription), show.legend =  FALSE) +
  geom_line(size = 0.5, show.legend =  FALSE) +
  geom_point(size = 2, show.legend =  FALSE) + 
  facet_wrap(~subscription) + 
  expand_limits(y = c(0, 50)) +
  geom_label(data = wk_df%>% filter(avg_dur_week == MAXc), aes(label = MAXc), label_size = 0.01,
    nudge_x = 0.25,
    vjust = -1,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = wk_df%>% filter(avg_dur_week == MAXm), aes(label = MAXm), label_size = 0.01,
    nudge_x = 0.25,
    vjust = -1,
    check_overlap = TRUE, show.legend =  FALSE) +
  geom_label(data = wk_df%>% filter(avg_dur_week == MINc), aes(label = MINc), label_size = 0.01,
    nudge_x = 0.25,
    vjust = 1,
    check_overlap = TRUE, show.legend =  FALSE) +
  geom_label(data = wk_df%>% filter(avg_dur_week == MINm), aes(label = MINm), label_size = 0.01,
    nudge_x = 0.25,
    vjust = 1.5,
    check_overlap = TRUE, show.legend =  FALSE) +
  geom_label(data = wk_df%>% filter(avg_dur_week == MAXc), aes(label = rd_week), label_size = 0.01,
    nudge_x = 0.25,
    nudge_y = 1,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = wk_df%>% filter(avg_dur_week == MINc), aes(label = rd_week), label_size = 0.01,
    nudge_x = 0.25,
    vjust = 2,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = wk_df%>% filter(avg_dur_week == MAXm), aes(label = rd_week), label_size = 0.01,
    nudge_x = 0.25,
    nudge_y = 1,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = wk_df%>% filter(avg_dur_week == MINm), aes(label = rd_week), label_size = 0.01,
    nudge_x = 0.25,
    vjust = 2.5,
    check_overlap = TRUE, show.legend =  FALSE) + 

labs(
    x = "Ride Week",
    y = "Average Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic"))
WAB

```

### Cyclistic Weekly Bike Total Usage For Annual Subscriptions(WBTA) (min)

```{r, fig.width = 5, fig.height=3}
# To split total weekly duration into annual memberships and  casual riders to get their individual min and max duration 

MAXm1 <- max(member_df %>% select(time))
MINm1 <- min(member_df %>% select(time))


MAXc1 <- max(casual_df %>% select(time))
MINc1 <- min(casual_df %>% select(time))


# to display plot
WBTA <-
  ggplot(member_df, aes(x = ryd_week, y = time)) +
  geom_bar(aes(fill = time),stat = "identity", show.legend =  FALSE, width = 1) +
  coord_flip() + theme(legend.position="none") +  
  scale_fill_gradient2(low ="orange",mid = "orange", high = "blue",space ="Lab", guide = "colourbar",
  aesthetics = "fill") +
  expand_limits(y = c(0, 125), x =c(0, 50)) +
  geom_label(data = member_df%>% filter(time == MAXm1), aes(label = MAXm), label_size = 0.01,
    hjust = -0.21,
    vjust = 0.2,
    check_overlap = TRUE, show.legend =  FALSE) +
  
  geom_label(data = member_df%>% filter(time == MINm1), aes(label = MINm), label_size = 0.01,
    nudge_x = 0.25,
    nudge_y = 0.05,
    check_overlap = TRUE, show.legend =  FALSE) +
   
  geom_text(data = member_df%>% filter(time == MAXm1), aes(label = rd_week), label_size = 0.01,
    hjust = -0.49,
    vjust = 1.5,
    check_overlap = TRUE, show.legend =  FALSE) +
  
 
  labs(
    x = "Ride Week",
    y = "Total Weekly Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic"))
WBTA

```

### Cyclistic Weekly Bike Total Usage For Casual Riders(WBTC)(min)

```{r, fig.width = 5, fig.height=3}
  
# to display plot
WBTC <- 
  ggplot(casual_df, aes(x = ryd_week, y = time)) +
  geom_bar(aes(fill = time),stat = "identity", show.legend =  FALSE, width = 1) +
  coord_flip() + theme(legend.position="none") + 
  scale_fill_gradient2(low ="#29C5F6",mid = "#FFCD91", high = "#58181F",space ="Lab", guide = "colourbar",
  aesthetics = "fill") +
  expand_limits(y = c(0, 295), x =c(0, 50)) +
  geom_label(data = casual_df%>% filter(time == MAXc1), aes(label = MAXc), label_size = 0.01,
    hjust = -0.25,
    vjust = 1.5,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = casual_df%>% filter(time == MINc1), aes(label = MINc), label_size = 0.01,
    nudge_x = 0.25,
    nudge_y = 1,
    check_overlap = TRUE, show.legend =  FALSE) +  
  geom_text(data = casual_df%>% filter(time == MAXc1), aes(label = rd_week), label_size = 0.01,
    hjust = -0.49,
    vjust = 0.5,
    check_overlap = TRUE, show.legend =  FALSE) +
  
 
  labs(
    x = "Ride Week",
    y = "Total Weekly Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic"))
WBTC

```

### Monthly Data Set

```{r}
mntly_df <- read.csv(url(mntly_data))
```

`head(mntly_df)`

```{r}
mntly_df$subscription <- mntly_df$member_casual%>%str_replace_all(c("member" = "annual subs", "casual" = "casual riders"))  
mntly_df$ryd_mnt <- mntly_df$ryd_month%>%str_replace_all(c("Jan" = "2021-01-01", "Feb" = "2021-02-01", "Mar" = "2021-03-01", "Apr" = "2021-04-01", "May" = "2021-05-01", "Jun" = "2021-06-01", "Jul" = "2021-07-01", "Aug" = "2021-08-01", "Sep" = "2021-09-01", "Oct" = "2021-10-01", "Nov" = "2021-11-01", "Dec" = "2021-12-01"))
#mntly_df <- arrange(mntly_df, ryd_mnt)  
mntly_df$ryd_mnt <- as.Date(mntly_df$ryd_mnt, format = "%Y-%m-%d")
  
mntly_df$time <- round((mntly_df$total_dur_mnt/ 10000), 2)

head(mntly_df)

```

| Ride Month | Member/Casual | Average Duration | Total Duration | subscription  | Ride Month(adjustted) |
|:-----------|:-----------|:-----------|:-----------|:-----------|:-----------|
| Jan        | member        | 12.90            | 1621055        | annual subs   | 2021-01-01            |
| Jan        | casual        | 26.78            | 2025831        | casual riders | 2021-01-01            |
| Feb        | member        | 13.78            | 1494415        | annual subs   | 2021-02-01            |
| Feb        | casual        | 28.48            | 2118231        | casual riders | 2021-02-01            |
| Mar        | casual        | 30.54            | 3848914        | casual riders | 2021-03-01            |
| Mar        | member        | 13.76            | 2264671        | annual subs   | 2021-03-01            |

: Cyclistic monthly bike usage

### Cyclistic Monthly Average Bike Usage(min)

```{r, fig.width = 8, fig.height=4}
manu_df <- mntly_df%>%filter(member_casual == "member")
Top_manu <- max(manu_df %>% select(avg_dur_mnt))

canu_df <-  mntly_df%>%filter(member_casual == "casual")
Top_canu <- max(canu_df%>% select(avg_dur_mnt))


ggplot(mntly_df, aes(x = ryd_mnt, y = avg_dur_mnt, color = subscription)) +
  geom_line(size = 0.5, show.legend =  FALSE) +
  geom_point(size = 2, show.legend =  FALSE) + 
  facet_wrap(~subscription) +
  expand_limits(y =c(10, 40)) +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b")  +  
  geom_label(data = manu_df%>% filter(avg_dur_mnt == Top_manu), aes(label = Top_manu), label_size = 0.01,
    hjust = 0.49,
    vjust = -1,
    check_overlap = TRUE, show.legend =  FALSE) + 
  geom_label(data = canu_df%>% filter(avg_dur_mnt == Top_canu), aes(label = Top_canu), label_size = 0.01,
    nudge_x = 0.25,
    nudge_y = 3.3,
    check_overlap = TRUE, show.legend =  FALSE) +  
  geom_text(data = manu_df%>% filter(avg_dur_mnt == Top_manu), aes(label = ryd_month), label_size = 0.01,
    hjust = 0.49,
    vjust = -0.5,
    check_overlap = TRUE, show.legend =  FALSE) +
  geom_text(data = canu_df%>% filter(avg_dur_mnt == Top_canu), aes(label = ryd_month), label_size = 0.01,
    hjust = 0.49,
    vjust = -0.5,
    check_overlap = TRUE, show.legend =  FALSE) +
 
  labs(
    x = "Ride Month",
    y = "Average Monthly Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic"))
```

### Cyclistic Monthly Bike Total Usage For Annual Subscriptions(min)

```{r, fig.width = 5, fig.height=4}
Topt_manu <- max(manu_df %>% select(time))


# to display plot
  ggplot(manu_df, aes(x = ryd_mnt, y = time)) +
  geom_bar(aes(fill = time),stat = "identity", show.legend =  FALSE, width = 26) +
  coord_flip() + theme(legend.position="none") +  
  scale_fill_gradient2(low ="orange",mid = "orange", high = "blue",space ="Lab", guide = "colourbar",
  aesthetics = "fill") +
  expand_limits(y = c(0, 450)) +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b") +               
  geom_label(data = manu_df%>% filter(time == Topt_manu), aes(label = Topt_manu), label_size = 0.01,
    hjust = -0.25,
    vjust = 0.1,
    check_overlap = TRUE, show.legend =  FALSE) +
  
  geom_text(data = manu_df%>% filter(time == Topt_manu), aes(label = ryd_month), label_size = 0.01,
    hjust = -0.9,
    vjust =1.6,
    check_overlap = TRUE, show.legend =  FALSE) +
  
 
  labs(
    x = "Ride Month",
    y = "Total Monthly Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic")) 
 

```

### Cyclistic Monthly Bike Total Usage For Casual Riders(min)

```{r, fig.width = 6, fig.height=4}
Topt_canu <- max(canu_df %>% select(time))


# to display plot
  ggplot(canu_df, aes(x = ryd_mnt, y = time)) +
  geom_bar(aes(fill = time),stat = "identity", show.legend =  FALSE, width = 26) +
  coord_flip() + theme(legend.position="none") +  
  scale_fill_gradient2(low ="#29C5F6",mid = "#FFCD91", high = "#58181F",space ="Lab", guide = "colourbar",
  aesthetics = "fill") +
  expand_limits(y = c(0, 870)) +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b") +               
  geom_label(data = canu_df%>% filter(time == Topt_canu), aes(label = Topt_canu), label_size = 0.01,
    hjust = -0.29,
    vjust = 0.05,
    check_overlap = TRUE, show.legend =  FALSE) +
  
  geom_text(data = canu_df%>% filter(time == Topt_canu), aes(label = ryd_month), label_size = 0.01,
    hjust = -1.6,
    vjust = 1.7,
    check_overlap = TRUE, show.legend =  FALSE) +
  
 
  labs(
    x = "Ride Month",
    y = "Total Monthly Duration (min)") +
  theme(
    axis.title.x = element_text(color = "#0099f9", size = 16, face = "bold"),
    axis.title.y = element_text(color = "#0099f9", size = 16, face = "italic")) 
 
```

### Year's Data Set

```{r}
yrs_df <- read.csv(url(years_data))
```

```{r, fig.width = 8}
yrs_df
```

| Member/Casual | Average Duration | Total Duration |
|---------------|------------------|----------------|
| member        | 13.31            | 33229048       |
| Casual        | 27.78            | 56000320       |

: Cyclistic's 2021 total bike usage

### [Key findings]{.underline}

casual riders use Cyclistic bikes more than Annual subscribers in terms of average duration. where casual riders used the Cyclistic bikes for 27.78 minutes per year on the average, annual subscribers used Cyclistic bikes for 13.31 minutes on the average as at 31 December 2021
